---
title: "biol499script4_h2"
format: html
editor: visual
date-modified: last-modified
date: today
author: kwu


toc: true
toc_depth: 3
---

# Description

Univariate analysis of biocrust x depth; I conducted a one-way anova on the 8 experimental groups including

# Additional notes

1.  Since it is an unbalanced design (no surface control), we cannot cross microsite and depth; therefore we must use a one-way anova on 8 experimental groups
2.  There is separate script to run anovas on just the groups; eg. 3 depths; 3 microsites

### Load libraries

```{r}

# stats
library(glmmTMB) # zero inflated data
library(lme4) # mixed-effects models
library(lmerTest) # provide p values for mixed-effect models
library(dplyr) # summarizing and tranform tables
library(DHARMa) # testing for zero inflation
library(tidyr)
library(car)

# visualization
library(ggplot2)
library(emmeans)
library(multcomp)
library(multcompView)
library(here)
```

### find wd using here
```{r}

gc <- read.csv(here("5_outputtables", "gc_wcategories.csv"))

```

### Prep data

```{r}
gc_clean <- gc %>%
  filter(microsite_condition != "double_control") %>%
  mutate(
    surface_depth = case_match(surface_depth, 
                               "10-May" ~ "5-10", 
                               .default = surface_depth),
    surface_depth = factor(surface_depth, levels = c("surface", "0-5", "5-10")),
    microsite_condition = factor(microsite_condition, 
                                 levels = c("biocrust", "dung", "control")),
    site = as.factor(site)
  )

# create master 160 list
all_trays <- expand_grid(
  site = factor(1:20),
  combo = c("biocrust_surface", "biocrust_0-5", "biocrust_5-10",
            "dung_surface", "dung_0-5", "dung_5-10",
            "control_0-5", "control_5-10")
) %>%
  separate(combo, into = c("microsite_condition", "surface_depth"), sep = "_")
```

# STATS

## ABUNDANCE

### Clean data

```{r}
# not normalizing abundance by volume (15.71) bc surface treatments are not standarized ----
# aggregate seedling counts by trays
tray_counts <- gc_clean %>%
  group_by(Tray.., site, microsite_condition, surface_depth) %>%
  summarise(total_seeds = n(), .groups = "drop") #drop grouping structure

tray_normal <- all_trays %>%
  left_join(tray_counts, by = c("site", "microsite_condition", "surface_depth")) %>%
  mutate(
    tray_id = row_number(),
    total_seeds = replace_na(total_seeds, 0), # assign 0 to trays with no seedlings
    treatment = paste(microsite_condition, surface_depth, sep = "_")
)

# set factors
treatment_levels <- c(
  "control_surface", "biocrust_surface", "dung_surface",
  "control_0-5",     "biocrust_0-5",     "dung_0-5",
  "control_5-10",    "biocrust_5-10",    "dung_5-10"
)

tray_normal <- tray_normal %>%
  mutate(
    treatment = factor(treatment, levels = treatment_levels),
    site = factor(site)
  )
```

### Run ANOVA/glmm

```{r}
#attempt 1: failed assumptions : glmer-----
#model_abundance <- glmer(seedpercm3 ~ microsite_condition * surface_depth + (1 | site), 
                        #data = tray_normal)

#anova(model_abundance) # anova table
#attempt 2: glmmTMB-----this model meets all assumptions----
#model_abundance <- glmmTMB(seedpercm3 ~ microsite_condition * surface_depth + (1|site), 
                                #data = tray_normal,
                                #family = tweedie())
#attempt 3: 
# count data; non-normal data, treatment = predictor, site as random effect
# glmmTMB for zero-inflated; Should i use?
model_abundance <- glmer(total_seeds ~ treatment + (1 | site), 
                         data = tray_normal,
                         family = poisson())

Anova(model_abundance, type = "III") # is this line of code okay? random effect model

```

### test assumptions

```{r}
# STEP 3: test assumptions ----
# check normality = right skewed
hist(tray_normal$total_seeds)

simulationOutput <- simulateResiduals(fittedModel = model_abundance)
plot(simulationOutput) # ks test(poor goodness of fit) and levene test(non-equal variance) is significant 
testOutliers(simulationOutput) #outlier test - not significant
testDispersion(simulationOutput)

```

### graph 1

```{r}

# STEP 1: get letters from the poisson count model
emm_counts <- emmeans(model_abundance, ~ treatment, type = "response")
emm_letters <- as.data.frame(cld(emm_counts, Letters = letters, adjust = "tukey")) %>%
  mutate(.group = trimws(.group)) %>%
  # IMPORTANT: Turn "biocrust_surface" back into two columns for the plot
  separate(treatment, into = c("microsite_condition", "surface_depth"), sep = "_") %>%
  mutate(
    surface_depth = factor(surface_depth, levels = c("surface", "0-5", "5-10")),
    microsite_condition = factor(microsite_condition, levels = c("biocrust", "dung", "control"))
  )

# STEP 2: calculate label positions
letter_pos <- tray_normal %>%
  group_by(microsite_condition, surface_depth) %>%
  summarise(y_pos = max(total_seeds, na.rm = TRUE) + 2, .groups = "drop") 

# Join letters to their positions
emm_letters <- left_join(emm_letters, letter_pos, by = c("microsite_condition", "surface_depth"))

depth_order <- c("surface", "0-5", "5-10")
micro_order <- c("dung","biocrust", "control")

# STEP 3: create the plot 
ggplot(tray_normal, aes(x = microsite_condition, y = total_seeds, fill = microsite_condition)) +
  # whiskers
  stat_boxplot(geom = "errorbar", width = 0.25) +
  
  # boxplot
  geom_boxplot(outlier.shape = 21, width = 0.6, alpha = 0.8) +
  
  # letter
  geom_text(data = emm_letters, 
            aes(x = microsite_condition, y = y_pos, label = .group),
            size = 5, fontface = "bold", vjust = 1) +
  
  # faceting
  facet_wrap(~ surface_depth, nrow = 1,
             labeller = labeller(surface_depth = c("surface" = "Surface", 
                                                   "0-5" = "0-5 cm", 
                                                   "5-10" = "5-10 cm"))) +
  
  # color
  scale_fill_manual(values = c("biocrust" = "#8B9E6B", 
                               "dung" = "#C4944A", 
                               "control" = "#D4C09A"),
                    labels = c("Dung", "Biocrust", "Control")) +
  
  # label
  labs(x = NULL, 
       y = "Seedling Count (per tray)", 
       fill = "Microsite") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    axis.text.x = element_blank(),   # Hides labels since colors/legend tell the story
    axis.ticks.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    text = element_text(size = 12)
  )
```

## RICHNESS

### Clean

```{r}
richness_counts <- gc_clean %>%
  group_by(site, microsite_condition, surface_depth) %>%
  summarise(richness = n_distinct(Species), .groups = "drop")

tray_richness <- all_trays %>%
  left_join(richness_counts, by = c("site", "microsite_condition", "surface_depth")) %>%
  mutate(
    richness = replace_na(richness, 0),
    treatment = factor(paste(microsite_condition, surface_depth, sep = "_"), 
                       levels = treatment_levels)
  )

```

### Run model

```{r}
# Run model with combined treatment
model_richness <- glmer(richness ~ treatment + (1|site), 
                          data = tray_richness,
                          family = poisson)

Anova(model_richness, type = "III")
```

### Test assumptions ----

```{r}
sim_richness <- simulateResiduals(model_richness) # non significant/pass assumptions
plot(sim_richness)
```

### graph2

```{r}
# Richness visualization -----

# STEP 1: get sig letters for treatment model
# = combined 'treatment' variable as the predictor
emm_rich <- emmeans(model_richness, ~ treatment, type = "response")
emm_rich_letters <- as.data.frame(cld(emm_rich, Letters = letters, adjust = "tukey")) %>%
  mutate(.group = trimws(.group)) %>%
  # split treatment
  separate(treatment, into = c("microsite_condition", "surface_depth"), sep = "_") %>%
  mutate(
    surface_depth = factor(surface_depth, levels = c("surface", "0-5", "5-10")),
    microsite_condition = factor(microsite_condition, levels = c("biocrust", "dung", "control"))
  )

# STEP 2: calculate letter positions
letter_pos_rich <- tray_richness %>%
  group_by(microsite_condition, surface_depth) %>%
  summarise(y_pos = max(richness, na.rm = TRUE) + 0.8, .groups = "drop") # 0.8 to clear whiskers

emm_rich_letters <- left_join(emm_rich_letters, letter_pos_rich, 
                              by = c("microsite_condition", "surface_depth"))

depth_order <- c("surface", "0-5", "5-10")
micro_order <- c("dung", "biocrust", "control")

# STEP 3: plot richness
ggplot(tray_richness, aes(x = microsite_condition, y = richness, fill = microsite_condition)) +
  # bars
  stat_boxplot(geom = "errorbar", width = 0.25) +
  
  # boxplot
  geom_boxplot(outlier.shape = 21, width = 0.6, alpha = 0.8) +
  
  # letters
  geom_text(data = emm_rich_letters, 
            aes(x = microsite_condition, y = y_pos, label = .group),
            size = 5, fontface = "bold", vjust = 1) +
  
  # subplots
  facet_wrap(~ surface_depth, nrow = 1,
             labeller = labeller(surface_depth = c("surface" = "Surface",
                                                   "0-5"     = "0-5 cm",
                                                   "5-10"    = "5-10 cm"))) +
  
  # color
  scale_fill_manual(values = c("biocrust" = "#8B9E6B",
                               "dung"      = "#C4944A",
                               "control"   = "#D4C09A"),
                    labels = c("Dung", "Biocrust", "Control")) +
  
  labs(x = NULL, 
       y = "Number of Species", 
       fill = "Microsite") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    axis.text.x = element_blank(),   
    axis.ticks.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    text = element_text(size = 12)
  )
```

## EVENNESS

### Calculate shannon h for pielou’s j (evenness) ----

```{r}
evenness_calc <- gc_clean %>%
  # grouping 
  group_by(site, microsite_condition, surface_depth, Species) %>% 
  # identifying species 
  summarise(sp_count = n(), .groups = "drop_last") %>%
  # calculates relative abundance
  mutate(
    p_i = sp_count / sum(sp_count), 
    # natural log
    ln_p_i = p_i * log(p_i)
  ) %>%
  summarise(
    # shannon diversity
    shannon_h = -sum(ln_p_i),
    # richness = tray_richness$richness
    richness_val = n(), 
    .groups = "drop"
  ) %>%
  mutate(
    # Pielou's evenness = H′/ln(S)
    # evenness requires at least 2 species to be mathematically defined
    evenness = if_else(richness_val > 1, shannon_h / log(richness_val), NA_real_)
  )
```

```{r}
# STEP 2: join with master list to include empty trays ----
tray_evenness <- all_trays %>%
  left_join(evenness_calc, by = c("site", "microsite_condition", "surface_depth")) %>%
  mutate(
    # Create the combined treatment factor
    treatment = factor(paste(microsite_condition, surface_depth, sep = "_"), 
                       levels = treatment_levels),
    # Beta distribution cannot handle exact 0 or 1.
    # Trays with 0 or 1 species result in NA
    evenness_raw = case_when(
      evenness <= 0 ~ 0.001,
      evenness >= 1 ~ 0.999,
      TRUE ~ evenness
    )
  )
```

### Run Model

```{r}
# use dispformula to fix the Levene test by modeling variance per microsite
model_evenness <- glmmTMB(evenness_raw ~ treatment + (1 | site), 
                          dispformula = ~microsite_condition, 
                          data = tray_evenness, 
                          family = beta_family())

Anova(model_evenness, type = "III")
```

### test assumptions

```{r}
sim_evenness <- simulateResiduals(model_evenness)
plot(sim_evenness) # good enough
```

### graph3

```{r}
# STEP 1: Get significance letters from the treatment-based model
emm_even <- emmeans(model_evenness, ~ treatment, type = "response")
emm_even_letters <- as.data.frame(cld(emm_even, Letters = letters, adjust = "tukey")) %>%
  mutate(.group = trimws(.group)) %>%
  # split the treatment back into microsite and depth
  separate(treatment, into = c("microsite_condition", "surface_depth"), sep = "_") %>%
  mutate(
    surface_depth = factor(surface_depth, levels = c("surface", "0-5", "5-10")),
    microsite_condition = factor(microsite_condition, levels = c("biocrust", "dung", "control"))
  )

# STEP 2: calculate letter positions
letter_pos_even <- tray_evenness %>%
  filter(!is.na(evenness)) %>%
  group_by(microsite_condition, surface_depth) %>%
  # position letters above max value
  summarise(y_pos = max(evenness, na.rm = TRUE) + 0.05, .groups = "drop") 

emm_even_letters <- left_join(emm_even_letters, letter_pos_even, 
                              by = c("microsite_condition", "surface_depth"))

# STEP 3: plot evenness
ggplot(tray_evenness %>% filter(!is.na(evenness)), 
       aes(x = microsite_condition, y = evenness, fill = microsite_condition)) +
  # Whiskers
  stat_boxplot(geom = "errorbar", width = 0.25) +
  # Main Boxplot
  geom_boxplot(outlier.shape = 21, width = 0.6, alpha = 0.8) +
  # Letters
  geom_text(data = emm_even_letters, 
            aes(x = microsite_condition, y = y_pos, label = .group),
            size = 4.5, fontface = "bold", vjust = 0) +
  # Faceting
  facet_wrap(~ surface_depth, nrow = 1,
             labeller = labeller(surface_depth = c(
               "surface" = "Surface", 
               "0-5" = "0-5 cm", 
               "5-10" = "5-10 cm"))) +
  scale_fill_manual(values = c("biocrust" = "#8B9E6B", 
                               "dung" = "#C4944A", 
                               "control" = "#D4C09A"),
                    labels = c("Biocrust", "Control", "Dung")) +
  # keeps scale clean
  scale_y_continuous(limits = c(0, 1.15), breaks = seq(0, 1, 0.2)) +
  labs(x = NULL, y = "Pielou's Evenness (J)", fill = "Microsite") +
  theme_classic() +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "bottom")

# Pielou's values go from 0-1. 
# Close to 1 means high evenness, ie there is no one dominant species
# Close to 0 means dominance by a single species
```
