---
title: "biol499script1_cleandescript"
format: html
editor: source
---

#load libraries
```{r}
# load packages
library(dplyr)
library(tidyverse)

```

# read and clean data
```{r}
# read data
df <- read.csv(here("2_data_raw", "gc_seedlingdata.csv")) #load data

# PART 1: clean data ----
# keep only the first 3 columns 
df <- df %>% 
  dplyr::select(1:3)

# Split the sample column into three columns and deal with double control 
df_cleaned <- df %>%
  separate("Sample", 
           into = c("site", "microsite_condition", "surface_depth"), 
           sep = "-", 
           remove = FALSE) %>% #keep original
  mutate(
    double_control = is.na(microsite_condition),
    microsite_condition = if_else(double_control, "double_control", microsite_condition),
    surface_depth = if_else(double_control, "double_control", surface_depth)
  )
    
df_cleannames <- df_cleaned %>%
    filter(microsite_condition != "double_control") %>%
    mutate(
    #recode microsite_condition with full names
    microsite_condition = case_when(
      microsite_condition == "D" ~ "dung",
      microsite_condition == "B" ~ "biocrust",
      microsite_condition == "C" ~ "control",
      TRUE ~ microsite_condition
    ),
    surface_depth = case_when(
      surface_depth == "S" ~ "surface",
      surface_depth == "0" ~ "0-5",
      surface_depth == "5" ~ "5-10",
      TRUE ~ surface_depth
    )
  ) 

write.csv(df_cleannames, here("5_outputtables", "gc_wcategories.csv"))
```

# descriptive analysis 

Remove control

# 1. number of species + count
```{r}
speciescount <- df_cleannames %>%
  group_by(Species) %>%
  summarise (count = n())

print(speciescount)

```

# 2. count per site
```{r}
sitecount <- df_cleannames %>%
  group_by(site) %>%
  summarise(
    count = n()
  )

print(sitecount)
```

# 3. gc sample count
```{r}
gctraycount <- df_cleannames %>%
  group_by(Sample) %>%
  summarize(
    count = n()
  )

print(gctraycount)
```

# 4. group by condition
```{r}
intx_trays <- df_cleannames %>%
  group_by(microsite_condition,surface_depth) %>%
  summarize(
    tray_num = n_distinct(Sample), 
    seedling_count = n(), 
    species_count = n_distinct(Species)) %>%
  arrange(desc(tray_num))

print(intx_trays)

write.csv(intx_trays, here("5_outputtables", "intx_trays.csv"))
```

# 5. Group by just microsites

```{r}
microsites <- df_cleannames %>%
  group_by(microsite_condition) %>%
  summarize(
    seedling_count = n(), 
    species_count = n_distinct(Species), 
    tray_count = n_distinct(Sample)
  )

print(microsites)
```

# visualizations
# Species
general and group by microsite
make stacked bar plots for by microsite

## clean data for visualization
```{r}

#append habit and origin to species 
spp <- read.csv(here("2_data_raw", "gc_species.csv"))

sppdf <- df_cleannames %>%
  left_join(spp, by = "Species")

# count species per microsite
# new row for distinct species for each microsite

sppms <- sppdf %>%
  group_by(microsite_condition, Species) %>%
  summarise(
    count = n(),
    .groups = "drop"
  )

```
### bar plot for nativity per microsite
```{r}

ggplot(sppms, aes(x = microsite_condition, y = count, fill =
                    Species)) +
  geom_col(color = "white", linewidth = 0.2) 
  
```

### general species count pie chart
```{r}
# organizing species

spp_pie <- sppdf %>%
  group_by(Species) %>%
  
  # figure out top #
  summarise(count = n(), groups = "drop") %>%
  
  mutate(top = fct_lump_n(Species, n = 8, w = count)) %>%
  
  # other
  group_by(top) %>%
  
  summarise(total_count = sum(count), .groups = "drop") %>%
  
  # order
  mutate(top = fct_reorder(top, total_count, .desc = TRUE),
         top = fct_relevel(top, "Other", after = Inf)) %>%
  arrange(top) %>%
  
  # calculate percentages for labels
  mutate(prop = total_count/sum(total_count),
         label = scales::percent(prop, accuracy =1))

# donut chart
ggplot(spp_pie, aes(x = 2, y = total_count, fill = top)) +
  
  # stack
  geom_col(color = "white") +
  # converts stack to donut
  coord_polar("y", start = 0) +
  
  # create hole
  xlim(0.5, 2.5) +
  
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), 
            size = 4, color = "white", fontface = "bold") +
  
  # color
  scale_fill_viridis_d(option = "turbo", name = "Species") +
  
  # theme
  theme_void() + 
  theme(legend.text = element_text(face = "italic"), # Italics for Latin names
        legend.title = element_text(face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16)) 
```

# nativity
```{r}
ggplot(sppdf, aes(x = microsite_condition, fill = Origin)) +
  geom_bar(position = "stack") + # geom_bar counts rows automatically
  scale_fill_manual(values = c("Native" = "#4E79A7", "Exotic" = "#E15759")) +
  theme_minimal()
```
# habit
```{r}
ggplot(sppdf, aes(x = microsite_condition, fill = Habit)) +
  geom_bar(position = "stack") + # geom_bar counts rows automatically
  scale_fill_manual(values = c("Forb" = "#4E79A7", "Graminoid" = "#E15759")) +
  theme_minimal()
```

# make sankey for connecting above to below
```{r}